Combinatorial Mixture Models for Single--Cell Assays with Application to Vaccine Studies

### 2012-07-25 MIMOSA.tex
Updated FDR curves for simulations and updated results text.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\title{Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies}
\author[1]{Greg Finak}
\author[2]{SC De Rosa}
\author[3]{Mario Roederer}
\author[1]{Raphael Gottardo}

@@

\begin{abstract}
Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on single-cell assays to provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
Here we present a Bayesian hierarchical framework based on beta-binomial mixture models for testing for differential marker expression using such single-cell assays. Our model allows the inference to be subject specific, while borrowing strength across subjects\footnote{Should we use subject, indivuduals or something else? What about animal studies?} through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental immunological assays measuring proteins or genes at the single-cell level, we find that our method has higher sensitivity and specificity than alternative methods. Using a simulation study we show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a multinomial-Dirichlet model. 
\end{abstract}

\section{Introduction}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers). More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells. A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}. While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability (add references). This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data. Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general. For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations. For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others thresholds are given by the assay technology. This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single cell level. After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers. Using this matrix, one can form $2^K$ putative cell subsets. When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. A common statistical problem is, for a given marker combination, to identify individuals for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination). A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify individuals' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells. While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future \cite{McKinstry:2010ei}. These antigen-specific T-cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells. The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine. Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}. A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted. These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences. individuals who generate a response after stimulation are called \textit{responders} whereas individual that do not show any differences are called \textit{non-responders}. In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.

Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that. Added ref for the HVTN054 trial and the ICS standardization paper from Horton})~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej}. 
All of these methods analyze each individual separately, and no information is shared across individuals even though one could expect some similarities across responders (\textit{resp.} non-responders). In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here? Let's update this sentence later}.

The framework developed in this paper addresses these issues explicitly. In our model, cell counts are modeled via a binomial (or multinomial in the multivariate case) distribution and information is shared across individuals by mean of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood. In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across individuals. This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single-cell\footnote{Let's replace all single cell by single-cell} assays, and increase sensitivity and specificity when detecting responders. Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\cite{Nason:2006dx}.

@@

\section{Results}
In this section, we apply our MIMOSA model to the data described in Section \ref{sec:data}, and present the results of a simulation study based on the ICS data.
%The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation. The unconstrained model was applied to Fluidigm single--cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single--cells. We also performed simulation studies to assess the performance of the constrained and unconstrained models in a univariate and multivariate settings.

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, ranked log fold change\footnote{Why ranked?}, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+?}. The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A. 
In addition, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B). These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figure~\ref{fig:HVTN065Results}) \footnote{Perhaps we could only keep the Env stimulation}.

\subsection{Single-cell gene expression}
We applied the MIMOSA model to Fluidigm single-cell gene expression. This time we used the two-sided MIMOSA model because genes could be up or down regulated upon stimulation. In order to detect stimulation specific changes of expression, our model was fit to each gene within each stimulation. The results presented in Figure~\ref{fig:fluidigm} A-C show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-individual variability (Figure~\ref{fig:fluidigm}). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions and the posterior log fold changes of proportions (Figure~\ref{fig:fluidigm} B,C). A similar analysis using a two-sided Fisher's exact test (Figure xx) does not reveal any stimulation specific changes of expression.\footnote{Perhaps it would be good to compare the number of differentially expressed genes}.

@@

We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations. Using hyper parameters estimated from a one-sided MIMOSA model fit to data from Gag1-stimulated, CD4-positive, IL2-expressing T-cells from day 28 post--vaccination of the HVTN054 trial~\cite{Peiperl:2010ej}. We simulated data from this constrained model with 500 observations, a response rate of 60\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, with ten independent realizations of data for each $N$. The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and ranked log fold change. This procedure was repeated for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A-D). In addition, the nominal \textit{vs.} observed FDR\footnote{Need to define this earlier} was also examined to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} E-F).

For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A, D). Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test (Figure~\ref{fig:simulations} E, F).
%Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Supplementary Figure~\ref{fig:simulations_trunc}). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.

\section{Differential expression across marker combinations}
Our beta-binomial model described in Section \ref{sec:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations. As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.\footnote{At some point we should say that such multivariate models are not appropriate for one sided models.} 
%Table~\ref{tab:multdir} gives an illustration of the counts with two putative markers A and B.
\subsection{Model}

@@

\subsection{Polyfunctionality in Fluidigm Single Cell Gene Expression Data}
As a proof-of-concept, we fitted our multivariate MIMOSA model looking at two specific genes in the Fluidigm data, namely BIRC3 and CCL5. In this case, $K=2$ and we have four possible combinations. 
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B). Only CCL5 positive cells express BIRC3, and the expression of BIRC3 increases upon stimulation. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible poly-functional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases. There is no difference in the marginal counts for any sample (Figure~\ref{fig:polyfunctionality} C). In contrast, multivariate MIMOSA tests all cell subpopulations simultaneously, and identifies a significant difference between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the individual combinations, which requires multiplicity adjustment and a potential loss of power.

@@

\section{Discussion}
Experimentalists have already access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few. Single cell assays will become even more routine once sequencing at the single-cell level becomes practical (Ref to RNA-Seq single cell). As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important. Current approaches for single-cell assays are for the most part simplistic (t-test, $\chi^2$ test, Fisher's exact test), and resulting inference can be quite unreliable especially when the cell counts are small. Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}. In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single cell assay.

@@

When changes are two-sided, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C). In this case, the use of multivariate model, as our Dirichlet-multinomial model, will become important to detect differential marker expression. Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single-cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B). Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model. The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$. As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters. A solution would be to explore alternative model parametrizations that could be used to reduce the number of required parameters. For example, one could assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$. As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others. More exploratory work will need to be done in this area once high dimensional single cell level data with large number of samples become available.

@@

\section*{Supplementary Information}
\subsection{HVTN065 and HVTN054 Vaccine Trial ICS Data Description}
\label{supp:statpublished}
HVTN065 is a phase 1 (safety and immunogenicity) trial of GeoVax HIV/AIDS DNA and MVA vaccine in 120 individuals (100 vaccinees, 20 placebo recipients, parts A and B). CD4 and CD8 T--cell epitope specific immune responses were measured via the ICS assay. Other humoral and cellular immune responses were measured via ELISA, and neutralizing antibody assays. Cytokines measured in the ICS assay included IFNg, TNFa, IL2, and IL4, and antigens included three Env, three Gag, and three Pol peptide pools. Results of the trial have been published~\citep{Goonetilleke:2006jk}.

>>>>

\title{Combinatorial Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
\author[1]{Greg Finak}
\author[2]{SC De Rosa}
\author[3]{Mario Roederer}
\author[1]{Raphael Gottardo}

@@

\begin{abstract}
Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single--cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on single-cell assays to provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
Here we present a Bayesian hierarchical framework based on beta-binomial mixture models for testing for differential marker expression using such single-cell assays. Our model allows the inference specific to each observation, while borrowing strength across observations through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental immunological assays measuring proteins or genes at the single-cell level, we find that our method has higher sensitivity and specificity than alternative methods. Using a simulation study we show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a multinomial-Dirichlet model. 
\end{abstract}

\section{Introduction}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single--cell measurement of multiple surface and intracellular markers (up to 17 markers). More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells. A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}. While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability (add references). This heterogeneity may carry important information, thus single--cell expression data should not be analyzed in the same fashion as cell-population level data. Special treatment of single--cell level data, which preserves information about population heterogeneity, is warranted in general. For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations. For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others thresholds are given by the assay technology. This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single--cell level. After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers. Using this matrix, one can form $2^K$ putative cell subsets. When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. A common statistical problem is, for a given marker combination, to identify individuals for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination). A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify individuals' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells. While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future \cite{McKinstry:2010ei}. These antigen-specific T-cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells. The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine. Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}. A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted. These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences. individuals who generate a response after stimulation are called \textit{responders} whereas individual that do not show any differences are called \textit{non-responders}. In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.

Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}. 
All of these methods test observations separately, and no information is shared across observations even though one could expect some similarities across responders (\textit{resp.} non-responders). %In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here? Let's update this sentence later}.

The framework developed in this paper addresses these issues explicitly. In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across individuals by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood. In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across individuals. This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders. Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\cite{Nason:2006dx}.

@@

\section{Results}
In this section, we apply our MIMOSA model to the data described in Section \ref{sec:data}, and present the results of a simulation study based on the ICS data. The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold--change by ROC (receiver operator characteristic) curve analysis and by comparing the observed FDR (false discovery rate) against the nominal FDR for each data set.

%The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation. The unconstrained model was applied to Fluidigm single--cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single--cells. We also performed simulation studies to assess the performance of the constrained and unconstrained models in a univariate and multivariate settings.

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold--change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+? Didn't look at CD8's}. The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold--change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A. 
In addition, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B). These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figure~\ref{fig:HVTN065Results}).

\subsection{Single-cell gene expression}
We applied the MIMOSA model to Fluidigm single-cell gene expression. This time we used the two-sided MIMOSA model because genes could be up or down regulated upon stimulation. In order to detect stimulation specific changes of expression, our model was fit to each gene within each stimulation. The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-individual variability (Figure~\ref{fig:fluidigm} A,B). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B). A similar analysis using a two-sided Fisher's exact test and clustering the signed q--values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation--specific patterns. Comparing the size of the gene sets, Fisher's exact test identified 47 significant genes while MIMOSA identified 50 significant genes, with 39 genes identified in common between the two methods.

@@

We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations. Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ENV--1--stimulated, CD4-positive, IL2-expressing T-cells from the primary immunogenicity time point of the HVTN065 trial~. We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000, and 50,000 events, with ten independent realizations of data for each $N$. The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold--change. This procedure was repeated for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A-D). In addition, the nominal \textit{vs.} observed FDR\footnote{Need to define this earlier} was also examined to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} A,B).

For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A and Supplementary Figure~\ref{fig:simulations_trunc} A). Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations} B and Supplementary Figure~\ref{fig:simulations_trunc} B).
%Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Supplementary Figure~\ref{fig:simulations_trunc} C,D). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.

\section{Differential expression across marker combinations}
Our beta-binomial model described in Section \ref{sec:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations. As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
%Table~\ref{tab:multdir} gives an illustration of the counts with two putative markers A and B.
\subsection{Model}

@@

\subsection{Polyfunctionality in Fluidigm Single--Cell Gene Expression Data}
As a proof-of-concept, we fitted our multivariate MIMOSA model looking at two specific genes in the Fluidigm data, namely BIRC3 and CCL5. In this case, $K=2$ and we have four possible combinations. 
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B). Only CCL5 positive cells express BIRC3, and the expression of BIRC3 increases upon stimulation. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases. There is no difference in the marginal counts for any sample (Figure~\ref{fig:polyfunctionality} C). In contrast, multivariate MIMOSA tests all cell subpopulations simultaneously, and identifies a significant difference between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the individual combinations, which requires multiplicity adjustment and a potential loss of power.

@@

\section{Discussion}
Experimentalists have already access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few. Single--cell assays will become even more routine once sequencing at the single-cell level becomes practical (Ref to RNA-Seq single--cell). As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single--cell level is becoming increasingly important. Current approaches for single-cell assays are for the most part simplistic (t-test, $\chi^2$ test, Fisher's exact test), and resulting inference can be quite unreliable especially when the cell counts are small. Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}. In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single--cell assay.

@@

When changes are two-sided, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C). In this case, the use of multivariate model, as our Dirichlet-multinomial model, will become important to detect differential marker expression. Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single-cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B). Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model. The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$. As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters. A solution would be to explore alternative model parametrizations that could be used to reduce the number of required parameters. For example, one could assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$. As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others. More exploratory work will need to be done in this area once high dimensional single--cell level data with large number of samples become available.

@@

\section*{Supplementary Information}
\subsection{HVTN065 Vaccine Trial ICS Data Description}
\label{supp:statpublished}
HVTN065 is a phase 1 (safety and immunogenicity) trial of GeoVax HIV/AIDS DNA and MVA vaccine in 120 individuals (100 vaccinees, 20 placebo recipients, parts A and B). CD4 and CD8 T--cell epitope specific immune responses were measured via the ICS assay. Other humoral and cellular immune responses were measured via ELISA, and neutralizing antibody assays. Cytokines measured in the ICS assay included IFNg, TNFa, IL2, and IL4, and antigens included three Env, three Gag, and three Pol peptide pools. Results of the trial have been published~\citep{Goonetilleke:2006jk}.
